
\chapter*{Introduction}\addcontentsline{toc}{chapter}{Introduction}\markboth{Introduction}{Introduction}
\setcounter{page}{1}

% Pohádka
% Nepřehánět to s definicemi a citacemi -> spíše teoretická část
% Proč to dělám, význam, návaznost (PyRigi)
% Jak jsem si to vybral
% prodat drony
% Cíle
% - část úvodu, nebo vlastní kapitola
% Obsah jen stručně

A core concept of Rigidity Theory is a \emph{framework} which is
a simple graph~\(G\) with its realization
into a \(d\)-dimensional space \(p: V(G) \to \R^d\).
A framework is \emph{\( d \)-flexible} if there exists
a continuous deformation of \( p \) such that
the distances between neighboring vertices are preserved during the transformation.
Otherwise, the framework is called \emph{\( d \)-rigid}.

It is known that for a graph either almost all realizations are \( d \)-flexible,
or almost all realizations are \( d \)-rigid~\cite{generically_rigid_graphs}.
A graph is \emph{(generically) \( d \)-rigid} if most of the realizations are \( d \)-rigid
and \emph{(generically) \( d \)-flexible} if most of the realizations are \( d \)-flexible.
%
As for a rigid graph,
there may be some flexible realizations,
and therefore we talk about paradoxical flexibility.

An illustrative application of the Rigidity Theory is a squadron of drones
where the drones can measure distance to their close neighbors.
Can we determine the locations of all the drones
up to translation and rotation of the whole squadron
just from such information?

For similar problems in the plane,
one would think that if the graph formed is rigid, the answer is yes, and
for flexible graphs the answer is no.
But as stated above, paradoxically, even a rigid graph can have a flexible realization,
and it may just happen that the drones form such a \( 2 \)-flexible framework.

In efforts to find such paradoxical realizations,
a new edge coloring was defined~\cite{legersky_original}.
A \emph{NAC-coloring} (where NAC stands for ``no almost cycle'')
is an~edge coloring of a graph by \( \red \) and \( \blue \)
such that both the colors are used and there is no almost cycle formed.
An \emph{almost cycle} is a cycle in the colored graph such that exactly one
edge of the cycle is colored \( \red \) or exactly one edge is colored \( \blue \).
Such coloring exists if and only if the graph has a flexible realization~\cite{legersky_original}.
This provides a simple but strong tool to decide whenever a graph has
a flexible realization even if it is (generically) rigid.
As shown later in the thesis, one can check in polynomial time
if a given coloring is a NAC-coloring.
Unfortunately, it is NP-complete to decide if a graph has any NAC-coloring~\cite{np_complete}.

\begin{figure}[ht]
	\centering
	\begin{tikzpicture}[rotate=90,scale=1.5]
		\node[vertex] (a) at (0,0) {};
		\node[vertex] (b) at (1,0) {};
		\node[vertex] (c) at (0.5,0.5) {};
		\node[vertex] (d) at (0,1.5) {};
		\node[vertex] (e) at (1,1.5) {};
		\node[vertex] (f) at (0.5,1) {};
		\draw[bedge] (a)edge(b) (b)edge(c) (c)edge(a) (d)edge(e) (e)edge(f) (f)edge(d) ;
		\draw[redge] (a)edge(d) (b)edge(e) (c)edge(f);
	\end{tikzpicture}
	\qquad
	\qquad
	\begin{tikzpicture}[rotate=90,scale=1.5]
		\node[vertex] (a) at (0.00,0) {};
		\node[vertex] (b) at (1.00,0) {};
		\node[vertex] (c) at (0.50,0.5) {};
		\node[vertex] (d) at (0.25,1) {};
		\node[vertex] (e) at (1.25,1) {};
		\node[vertex] (f) at (0.75,1.5) {};
		\draw[edge] (a)edge(b) (b)edge(c) (c)edge(a) (d)edge(e) (e)edge(f) (f)edge(d) ;
		\draw[edge] (a)edge(d) (b)edge(e) (c)edge(f);
	\end{tikzpicture}
	\caption[Flexible realization of 3-prism]{The $3$-prism is generically $2$-rigid, but has a flexible realization (right).
		It has a unique NAC-coloring modulo swapping colors (left).}%
	\label{fig:3prism}
\end{figure}

A nice example for NAC-coloring showcase is the 3-prism graph in~\Cref{fig:3prism},
a graph formed from two triangles with three interconnecting edges.
Such a graph is rigid, still it has a NAC-coloring and flexible realizations.
The flexible realizations are all the realizations where
the triangles are identical except for translation.
It can be easily seen that there are
no other NAC-colorings except for the one, where the colors are swapped.

Among the goals of this thesis is to
%
study basics of Rigidity theory with the focus on existence of flexible realizations
and its combinatorial characterization using NAC-colorings.
%
In \Cref{chapter:np},
we show that the problem of deciding if a graph has a NAC-coloring
is NP-complete even on graphs with maximum degree five.
%
We provide a parametrized approach in \Cref{chapter:fpt}
where complexity can be reduced for graphs
with low internal structural complexity (treewidth).
%
The relation of NAC-colorings with stable cuts is presented in \Cref{chapter:stable_cuts},
and an algorithm for stable cut search on flexible graphs from~\cite{stable_cuts_legersky}
is described and implemented.
%
In \Cref{chapter:algo},
we provide an algorithm
that can find all the NAC-colorings of a graph.
Multiple heuristics are used to significantly
reduce the runtime of the algorithm.
%
We describe the implementation and
compare and evaluate the performance of the algorithm
compared to previous implementations,
and also we compare individual heuristics among each other
in \Cref{chapter:benchmarks}.
Multiple relevant graph classes are considered.

The code implemented in this thesis available at~\cite{my_code}
shall enrich PyRigi~\cite{pyrigi}
library used for research in Rigidity Theory.
Currently, there are no such algorithms implemented in PyRigi itself,
the only available implementation is rather naive in FlexRiLoG~\cite{flexrilog}.

In this thesis, the results shown in our previous paper~\cite{my_paper}
(done as part of VýLeT --- Student Summer Research Program)
are extended upon by introducing a fixed-parameter tractable algorithm,
discussion of relation to stable cuts,
and present other heuristic that do not perform as well.
Here, we also describe the remaining topics more in detail.

